FAIR Theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories

### 2024-09-13 manuscript.Rmd
update 2024-09-13 10:38:59.00593

The FAIR Guiding Principles (hereafter: FAIR principles) were initially established to make research data more Findable, Accessible, Interoperable and Reusable [REF].
Since they were first introduced, scholars have demonstrated the relevance of these criteria for making other digital research artefacts more open.
This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory as well.
We introduce "FAIR theory", a digital representation of theory that can complies with the FAIR principles.

<!-- QUOTES -->
Kurt Lewin: Nothing is as practical as a good theory
Paul Meehl: [Psychological theories are neither] refuted nor corroborated, but instead merely fade away as people lose interest [@meehlTheoreticalRisksTabular1978].
[Theories are] like toothbrushes — no self-respecting person wants to use anyone else’s [@mischelToothbrushProblem2008]
"Call Leon": When finding unexpected results in cognitive dissonance research, students were told to call Leon Festinger. He could explain why the study did not work, e.g. because you used the wrong font for the questionnaire.

<!-- END QUOTES -->

The "replication crisis" has prompted extensive reforms in social science [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].

Concern that undisclosed flexibility in analyses was to blame for the abundance of false-positive findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
But have these reforms met their goal?
Recent reviews show that most preregistered hypothesis tests are not supported [@scheelExcessPositiveResults2021],
which suggests that the replication crisis may have been symptomatic of an underlying, and more fundamental, "theory crisis".
If psychological theories are insufficiently precise to derive testable hypotheses - then preregistration and replication will only serve to highlight those shortcomings.

Scholars have raised concerns about the state of theory in social science for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
There is increasing consensus that one contributing cause is a lack of theory formalization: social scientific theories often lack the precision and clarity of theories in the physical sciences [@szollosiArrestedTheoryDevelopment2021].
A second issue, which received less attention, is the a lack of transparent and democratic scholarly communication about psychological theory.
The present paper seeks to address this issue by applying open science principles, for the first time, to psychological theory, and introducing the concept of *FAIR Theory*.

@@

Theories reflect scientists' understanding of phenomena.
Ideally, they are iteratively updated based on deductive testing and inductive theory construction.
This process is described in the *empirical cycle* [@degrootMethodologieGrondslagenVan1961],
a model that describes cumulative knowledge acquisition as a cyclical process with two phases (Figure \@ref(fig:figec)).
In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.

@@

In a progressive research program [@lakatosHistoryScienceIts1971],
the cycle is regularly completed to iteratively advance our understanding of the studied phenomena.

Contemporary psychological research falls short of this idealized model in several ways.
Firstly, because deductive research is overrepresented in the literature. 
According to one estimate, 89.6% of published studies tests hypotheses, which suggests that the literature is predominantly comprised of deductive research  [@kuhbergerPublicationBiasPsychology2014].
Closer examination reveals, however, that the link between theory and hypothesis is often tenuous [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
Only 15% of deductive studies even reference specific theories [@mcphetresDecadeTheoryReflected2021].
Together, these statistics suggest the paradoxical role of theory in contemporary psychology.
Papers set out to test hypotheses, but these are rarely derived from theory, and test results rarely contribute to the improvement of existing theories.
Existing theories either persist unchanged for decades, or are forgotten [REF Meehl].

@@

Merely publishing a theory does not make it open;
open theory should adhere to established open science standards.
The FAIR principles, initially introduced as a standard for open research data, have since been applied to other forms of digital scholarly output [e.g., software @lamprechtFAIRPrinciplesResearch2019].
We propose to apply the FAIR principles to digital representations of theory as well,
introducing a FAIR metadata format to represent (formal) theories.
The resulting theories are made *Findable* via a DOI,
are Accessible in a machine- and human-readable filetype,
Interoperable within the data analysis environment,
and Reusable in the practical and legal sense, using semantic versioning and permissive licensing.

@@

The FAIR Principles were devised to make scholarly data more findable, accessible, interoperable, and reusable. From their inception, these principles were developed with "other research resources" in mind. Since then, scholars have translated the FAIR principles to, e.g., research software [REF Lamprecht]. The present paper further extends the FAIR principles' definition to theory. Doing so requires first representing theory as a digital research artefact, and then making it compliant with the FAIR principles.

@@

Is it findable? There is no unified search engine for theory, and not even an agreed-upon search term (model, framework, etc are used interchangeably with theory). Some have sought to address findability through post-hoc curation of theories (e.g., theory mapping website, Borsboom's dictionary of psychological phenomena).

Is it accessible? Paywalls, partly implicit (Great Man Theorizing / Call Leon), certainly not accessible in the sense of theory being "top of mind", as researchers appear to prefer ad-hoc hypothesizing rather than deriving hypotheses from specific theories.

Is it interoperable? Theories rarely change,

>>>>

The FAIR Guiding Principles (hereafter: FAIR principles) were established to make research data more Findable, Accessible, Interoperable and Reusable [REF].
Since they were first introduced, scholars have demonstrated their relevance for making other digital research artefacts more open.
This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory as well.
To this end, we introduce the concept of "FAIR theory",
a digital representation of theory compliant with the FAIR principles.

<!-- QUOTES

Kurt Lewin: Nothing is as practical as a good theory
Paul Meehl: [Psychological theories are neither] refuted nor corroborated, but instead merely fade away as people lose interest [@meehlTheoreticalRisksTabular1978].
[Theories are] like toothbrushes — no self-respecting person wants to use anyone else’s [@mischelToothbrushProblem2008]
"Call Leon": When finding unexpected results in cognitive dissonance research, students were told to call Leon Festinger. He could explain why the study did not work, e.g. because you used the wrong font for the questionnaire.

END QUOTES -->

The "replication crisis" has prompted extensive reforms in social science [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
Concern that undisclosed flexibility in analyses was to blame for the abundance of false-positive findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
But have these reforms met their goal?
Recent reviews show that most preregistered hypothesis tests are not supported [@scheelExcessPositiveResults2021].
One plausible explanation is that the replication crisis is symptomatic of an underlying,
and more fundamental, "theory crisis".
If psychological theories are insufficiently precise to derive testable hypotheses,
or sufficiently vague to explain contradictory findings,
then preregistration and replication will only serve to highlight those shortcomings.

Scholars have raised concerns about the state of theory in social science for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
One contributing factor is a lack of theory formalization:
social scientific theories often lack the precision and clarity of theories in the physical sciences [@szollosiArrestedTheoryDevelopment2021].
A second factor, which received less attention, is the a lack of transparent and democratic scholarly communication about psychological theory.
The present paper seeks to advance transparent communication about theory by applying open science principles to psychological theory for the first time,
and introducing the concept of *FAIR Theory*.

@@

According to the *empirical cycle* [@degrootMethodologieGrondslagenVan1961],
a philosophical model of cumulative knowledge acquisition,
research ideally follows a cyclical process with two phases (Figure \@ref(fig:figec)).
In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.
In this model, theories are the vehicle of scientists' understanding of phenomena.
Ideally, they are iteratively updated based on deductive testing and inductive theory construction.

@@

In a progressive research program [@lakatosHistoryScienceIts1971],
the cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
There are clear indications that contemporary psychology falls short of this idealized model, however.
Firstly, because deductive research is over-represented in the literature. 
According to one estimate, 89.6% of published studies tests hypotheses,
which suggests that the literature is predominantly comprised of deductive research  [@kuhbergerPublicationBiasPsychology2014].
Closer examination reveals, however, that the link between theory and hypothesis is often tenuous [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
Only 15% of deductive studies reference theory at all [@mcphetresDecadeTheoryReflected2021].
This raises the question where the other hypotheses come from,
and what consequence their refutation would have for theory.
These statistics suggest that theory has an uncomfortable and paradoxical role in contemporary psychology:
The majority of papers ostensibly test hypotheses,
but these are rarely derived from theory,
and test results rarely contribute to the improvement of existing theories.
Consequently, theories either persist unchanged for decades, or are forgotten [REF Meehl].

@@

Merely publishing a theory does not make it open;
to be open, theory should adhere to established open science standards.
The FAIR principles, initially introduced as a standard for open research data, have since been applied to other forms of digital scholarly output [e.g., software @lamprechtFAIRPrinciplesResearch2019].
We propose to apply the FAIR principles to digital representations of theory as well,
introducing a FAIR metadata format to represent (formal) theories.
The resulting theories are made *Findable* via a DOI,
Accessible in a machine- and human-readable filetype,
Interoperable within the data analysis environment,
and Reusable in the practical and legal sense, so that they may be improved over time.

@@

The FAIR Principles were devised to make scholarly data more findable, accessible, interoperable, and reusable.
From their inception, these principles were developed with "other research resources" in mind.
Scholars have translated the FAIR principles to, e.g., research software [REF Lamprecht].
The present paper further extends the FAIR principles' definition to theory, see Table \@ref(tab:tabfair).

```{r tabfair}
tab_fair <- read.csv("fair_principles.csv", stringsAsFactors = FALSE)[, c("Criterion", "Original", "Theory", "Our.action")]
names(tab_fair) <- c("Criterion", "Original", "Theory", "Action")
papaja::apa_table(tab_fair, landscape = TRUE, align = "m{.1\\linewidth}m{.35\\linewidth}m{.35\\linewidth}m{.15\\linewidth}")
```

@@

While a comprehensive analysis of the present state of FAIR Theory in psychology is beyond the scope of the present paper,
we provide a narrative review of examples and counterexamples for each criterion.

One factor that contributes to poor findability of psychological theory is that the primary unit of dissemination and search in psychology is the academic paper.
A paper may contain multiple resources - including materials, data, code, and theory - but there is no unified search engine for theory,
or even an agreed-upon keyword (model, framework, etc are often used interchangeably with theory).
Modular publishing could ameliorate the Findability of psychological theory.
In this approach, distinct resources are individually published as citable academic output.
Such output can still be linked; for example, a FAIR theory could have an accompanying paper.
Another effort to improve theories' findability is post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and post many examples on their website.
Similarly, Borsboom and colleagues seek to establish a dictionary of psychological "phenomena" (which are not strictly theories, but are patterns reliably evidenced by data that theory should seek to explain).

With regard to Accessibility, publishing behind paywalls certainly has a negative impact.
Open Access publishing increases the accessibility of all academic output, including theory.
Another factor curtailing the accessibility of theory is ambiguity;
a lack of theory formalization introduces a dependency on the original author for clarification.
The discourse on "Great Man Theorizing" touches upon the problems this introduces [REF Guest et al moral theory].
For example, dependency on interpretation by the author creates a potential for gatekeeping - the author could insist that work requires their involvement, which violates checks and balances of scientific research.
Moreover, if a theory is refuted, its author could claim that the author of the refuting paper did not interpret the theory correctly.
To some extent, this relates to the problem of translation [REF Duhem]:
it is not possible to entirely formalize an idea to enable unambiguous interpretation.
Nonetheless, taking care to formalize a theory to the maximum extent possible advances Accessibility.

 <!-- (Great Man Theorizing / Call Leon), certainly not accessible in the sense of theory being "top of mind", as researchers appear to prefer ad-hoc hypothesizing rather than deriving hypotheses from specific theories. -->

The primary indication that psychological theory has limited interoperability ? Theories rarely change,

### 2024-09-13 manuscript.tex
update 2024-09-13 10:38:59.00593

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{keywords\newline\indent Word count: X}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={FAIR Theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories},
  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2, Andreas Brandmaier2, \& Felix Schönbrodt2},
  pdflang={en-EN},
  pdfkeywords={keywords},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{FAIR Theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories}
\author{Caspar J. Van Lissa\textsuperscript{1}, Aaron Peikert\textsuperscript{2}, Andreas Brandmaier\textsuperscript{2}, \& Felix Schönbrodt\textsuperscript{2}}
\date{}

@@

This is a preprint paper, generated from Git Commit \# fab0d274.

The authors made the following contributions. Caspar J. Van Lissa: Conceptualization, Formal Analysis, Funding acquisition, Methodology, Project administration, Software, Supervision, Writing -- original draft, Writing -- review \& editing; Aaron Peikert: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Andreas Brandmaier: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Felix Schönbrodt: Conceptualization, Writing -- review \& editing.

@@

\affiliation{\vspace{0.5cm}\textsuperscript{1} Tilburg University, dept. Methodology \& Statistics\\\textsuperscript{2} Other affiliations}

@@

Kurt Lewin: Nothing is as practical as a good theory
Paul Meehl: {[}Psychological theories are neither{]} refuted nor corroborated, but instead merely fade away as people lose interest (Meehl, 1978).
{[}Theories are{]} like toothbrushes --- no self-respecting person wants to use anyone else's (Mischel, 2008)
``Call Leon'': When finding unexpected results in cognitive dissonance research, students were told to call Leon Festinger. He could explain why the study did not work, e.g.~because you used the wrong font for the questionnaire.

The ``replication crisis'' has prompted extensive reforms in social science (Lavelle, 2021; Scheel, 2022).
Concern that undisclosed flexibility in analyses was to blame for the abundance of false-positive findings led to widespread adoption of open science practices like preregistration and replication (Nosek et al., 2015).
But have these reforms met their goal?
Recent reviews show that most preregistered hypothesis tests are not supported (Scheel, Schijen, \& Lakens, 2021),
which suggests that the replication crisis may have been symptomatic of an underlying, and more fundamental, ``theory crisis''.
If psychological theories are insufficiently precise to derive testable hypotheses - then preregistration and replication will only serve to highlight those shortcomings.

Scholars have raised concerns about the state of theory in social science for nearly 50 years (Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
There is increasing consensus that one contributing cause is a lack of theory formalization: social scientific theories often lack the precision and clarity of theories in the physical sciences (Szollosi \& Donkin, 2021).
A second issue, which received less attention, is the a lack of transparent and democratic scholarly communication about psychological theory.
The present paper seeks to address this issue by applying open science principles, for the first time, to psychological theory, and introducing the concept of \emph{FAIR Theory}.

@@

Theories reflect scientists' understanding of phenomena.
Ideally, they are iteratively updated based on deductive testing and inductive theory construction.
This process is described in the \emph{empirical cycle} (de Groot, 1961),
a model that describes cumulative knowledge acquisition as a cyclical process with two phases (Figure \ref{fig:figec}).
In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.

@@

In a progressive research program (Lakatos, 1971),
the cycle is regularly completed to iteratively advance our understanding of the studied phenomena.

When comparing common practice in contemporary psychological research to this idealized model, at first glance,
it appears that deductive research is overrepresented in the literature.
An overwhelming 89.6\% of published studies ostensibly tests hypotheses (Kühberger, Fritz, \& Scherndl, 2014).
Closer examination reveals, however, that the link between theory and hypothesis is often tenuous (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
Only 15\% of deductive studies even reference specific theories (McPhetres et al., 2021).
Together, these statistics suggest the paradoxical role of theory in contemporary psychology.
Papers set out to test hypotheses, but these are rarely derived from theory, and test results rarely contribute to the improvement of existing theories.
Existing theories either persist unchanged, or are forgotten {[}REF Meehl{]}.

@@

Merely publishing a theory does not make it open;
open theory should adhere to established open science standards.
The FAIR principles, initially introduced as a standard for open research data, have since been applied to other forms of digital scholarly output (e.g., software Lamprecht et al., 2019).
We propose to apply the FAIR principles to digital representations of theory as well,
introducing a FAIR metadata format to represent (formal) theories.
The resulting theories are made \emph{Findable} via a DOI,
are Accessible in a machine- and human-readable filetype,
Interoperable within the data analysis environment,
and Reusable in the practical and legal sense, using semantic versioning and permissive licensing.

@@

The FAIR Principles were devised to make scholarly data more findable, accessible, interoperable, and reusable. From their inception, these principles were developed with ``other research resources'' in mind. Since then, scholars have translated the FAIR principles to, e.g., research software {[}REF Lamprecht{]}. The present paper further extends the FAIR principles' definition to theory. Doing so requires first representing theory as a digital research artefact, and then making it compliant with the FAIR principles.

@@

Is it findable? There is no unified search engine for theory, and not even an agreed-upon search term (model, framework, etc are used interchangeably with theory). Some have sought to address findability through post-hoc curation of theories (e.g., theory mapping website, Borsboom's dictionary of psychological phenomena).

Is it accessible? Paywalls, partly implicit (Great Man Theorizing / Call Leon), certainly not accessible in the sense of theory being ``top of mind'', as researchers appear to prefer ad-hoc hypothesizing rather than deriving hypotheses from specific theories.

@@

Based on this example: \url{https://www.r-bloggers.com/2019/08/causal-inference-with-dags-in-r/}

@@

\bibitem[\citeproctext]{ref-mischelToothbrushProblem2008}
Mischel, W. (2008). The {Toothbrush Problem}. \emph{APS Observer}, \emph{21}. Retrieved from \url{https://www.psychologicalscience.org/observer/the-toothbrush-problem}

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 1808}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={FAIR Theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories},
  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2,3, Andreas M. Brandmaier2,3,4, \& Felix Schönbrodt5},
  pdflang={en-EN},
  pdfkeywords={meta theory, theory formation, cumulative science, formal models},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{FAIR Theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories}
\author{Caspar J. Van Lissa\textsuperscript{1}, Aaron Peikert\textsuperscript{2,3}, Andreas M. Brandmaier\textsuperscript{2,3,4}, \& Felix Schönbrodt\textsuperscript{5}}
\date{}

@@

This is a preprint paper, generated from Git Commit \# 50b18562.

The authors made the following contributions. Caspar J. Van Lissa: Conceptualization, Formal Analysis, Funding acquisition, Methodology, Project administration, Software, Supervision, Writing -- original draft, Writing -- review \& editing; Aaron Peikert: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Andreas M. Brandmaier: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Felix Schönbrodt: Conceptualization, Writing -- review \& editing.

@@

\affiliation{\vspace{0.5cm}\textsuperscript{1} Tilburg University, dept. Methodology \& Statistics\\\textsuperscript{2} Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany\\\textsuperscript{3} Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany\\\textsuperscript{4} Department of Psychology, MSB Medical School Berlin, Berlin, Germany\\\textsuperscript{5} Other affiliations}

@@

The FAIR Guiding Principles (hereafter: FAIR principles) were established to make research data more Findable, Accessible, Interoperable and Reusable {[}REF{]}.
Since they were first introduced, scholars have demonstrated their relevance for making other digital research artefacts more open.
This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory as well.
To this end, we introduce the concept of ``FAIR theory'',
a digital representation of theory compliant with the FAIR principles.

The ``replication crisis'' has prompted extensive reforms in social science (Lavelle, 2021; Scheel, 2022).
Concern that undisclosed flexibility in analyses was to blame for the abundance of false-positive findings led to widespread adoption of open science practices like preregistration and replication (Nosek et al., 2015).
But have these reforms met their goal?
Recent reviews show that most preregistered hypothesis tests are not supported (Scheel, Schijen, \& Lakens, 2021).
One plausible explanation is that the replication crisis is symptomatic of an underlying,
and more fundamental, ``theory crisis''.
If psychological theories are insufficiently precise to derive testable hypotheses,
or sufficiently vague to explain contradictory findings,
then preregistration and replication will only serve to highlight those shortcomings.

Scholars have raised concerns about the state of theory in social science for nearly 50 years (Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
One contributing factor is a lack of theory formalization:
social scientific theories often lack the precision and clarity of theories in the physical sciences (Szollosi \& Donkin, 2021).
A second factor, which received less attention, is the a lack of transparent and democratic scholarly communication about psychological theory.
The present paper seeks to advance transparent communication about theory by applying open science principles to psychological theory for the first time,
and introducing the concept of \emph{FAIR Theory}.

@@

According to the \emph{empirical cycle} (de Groot, 1961),
a philosophical model of cumulative knowledge acquisition,
research ideally follows a cyclical process with two phases (Figure \ref{fig:figec}).
In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.
In this model, theories are the vehicle of scientists' understanding of phenomena.
Ideally, they are iteratively updated based on deductive testing and inductive theory construction.

@@

In a progressive research program (Lakatos, 1971),
the cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
There are clear indications that contemporary psychology falls short of this idealized model, however.
Firstly, because deductive research is over-represented in the literature.
According to one estimate, 89.6\% of published studies tests hypotheses,
which suggests that the literature is predominantly comprised of deductive research (Kühberger, Fritz, \& Scherndl, 2014).
Closer examination reveals, however, that the link between theory and hypothesis is often tenuous (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
Only 15\% of deductive studies reference theory at all (McPhetres et al., 2021).
This raises the question where the other hypotheses come from,
and what consequence their refutation would have for theory.
These statistics suggest that theory has an uncomfortable and paradoxical role in contemporary psychology:
The majority of papers ostensibly test hypotheses,
but these are rarely derived from theory,
and test results rarely contribute to the improvement of existing theories.
Consequently, theories either persist unchanged for decades, or are forgotten {[}REF Meehl{]}.

@@

Merely publishing a theory does not make it open;
to be open, theory should adhere to established open science standards.
The FAIR principles, initially introduced as a standard for open research data, have since been applied to other forms of digital scholarly output (e.g., software Lamprecht et al., 2019).
We propose to apply the FAIR principles to digital representations of theory as well,
introducing a FAIR metadata format to represent (formal) theories.
The resulting theories are made \emph{Findable} via a DOI,
Accessible in a machine- and human-readable filetype,
Interoperable within the data analysis environment,
and Reusable in the practical and legal sense, so that they may be improved over time.

@@

The FAIR Principles were devised to make scholarly data more findable, accessible, interoperable, and reusable.
From their inception, these principles were developed with ``other research resources'' in mind.
Scholars have translated the FAIR principles to, e.g., research software {[}REF Lamprecht{]}.
The present paper further extends the FAIR principles' definition to theory, see Table \ref{tab:tabfair}.

\begin{lltable}

\begin{longtable}{m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
\caption{\label{tab:tabfair}}\\
\toprule
Criterion & \multicolumn{1}{c}{Original} & \multicolumn{1}{c}{Theory} & \multicolumn{1}{c}{Action}\\
\midrule
\endfirsthead
\caption*{\normalfont{Table \ref{tab:tabfair} continued}}\\
\toprule
Criterion & \multicolumn{1}{c}{Original} & \multicolumn{1}{c}{Theory} & \multicolumn{1}{c}{Action}\\
\midrule
\endhead
F1 & (Meta)data are assigned a globally unique and persistent identifier & Theory and its associated metadata has a global, unique and persistent identifier for each version (using semantic versioning) & Rephrased\\
F2 & Data are described with rich metadata & Theory is described with rich metadata & \textasciitilde{}same\\
F3 & Metadata clearly and explicitly include the identifier of the data it describes & Metadata clearly and explicitly include identifiers for all the versions of the theory it describes & \textasciitilde{}same\\
F4 & (Meta)data are registered or indexed in a searchable resource & Theory and its associated metadata are included in a searchable repository & NEEDS WORK! GitHub is indexed by Google I believe, but ideally, we'd like our theories to show up in Google Scholar, or even dedicated academic search enginges. Where could we put them to realize this?\\
A1 & (Meta)data are retrievable by their identifier using a standardized communications protocol & Theory and its associated metadata are accessible by their identifier using a standardized communications protocol & \textasciitilde{}same\\
A1.1 & The protocol is open, free, and universally implementable & The protocol is open, free, and universally implementable & Same\\
A1.2 & The protocol allows for an authentication and authorization procedure, where necessary & The protocol allows for an authentication and authorization procedure, where necessary & Do\ \ we need this?\\
A2 & Metadata are accessible, even when the data are no longer available & Theory metadata are accessible, even when the theory is no longer available & Do\ \ we need this?\\
I1 & (Meta)data use a formal, accessible, shared, and broadly applicable language for knowledge representation & Theory and its associated metadata use a formal, accessible, shared and broadly applicable language to facilitate machine readability and reuse & Rephrased\\
I2 & (Meta)data use vocabularies that follow FAIR principles & (Meta)data use vocabularies that follow FAIR principles & NEEDS WORK! I think this is where we explain the value of e.g. universal graph languages like Aaron and Max'\\
I2S.1 & - &  & \\
I2S.2 & - &  & \\
I3 & (Meta)data include qualified references to other (meta)data & (Meta)data includes qualified references to other (meta)data, including previous versions of the theory & Rephrased. I envision a LinkList-like structure where each theory version references its ancestor\\
I4S & - &  & Discard\\
R1 & (Meta)data are richly described with a plurality of accurate and relevant attributes & Theory and its associated metadata are richly described with a plurality of accurate and relevant attributes & Needs work. How do we envision this? Keywords? ISBN-like codes for variable type?\\
R1.1 & (Meta)data are released with a clear and accessible data usage license & (Meta)data are released with a clear and accessible license & Needs work: which license is good for theory?\\
R1.2 & (Meta)data are associated with detailed provenance & (Meta)data are associated with detailed provenance & Relates to I3; I think this point relates more to the theory's ancestry, and I3 relates to e.g. incorporating other theories within a theory (e.g., theory of measurement inside of structural theory)\\
R1.3 & (Meta)data meet domain-relevant community standards & Theory metadata and documentation meet domain-relevant community standards & These standards do not yet exist; we can take a first step towards developing them and recommend that this be an active area of development\\
\bottomrule
\end{longtable}

\end{lltable}

@@

While a comprehensive analysis of the present state of FAIR Theory in psychology is beyond the scope of the present paper,
we provide a narrative review of examples and counterexamples for each criterion.
One factor that contributes to poor findability of psychological theory is that the primary unit of dissemination and search in psychology is the academic paper.
A paper may contain multiple resources - including materials, data, code, and theory - but there is no unified search engine for theory,
or even an agreed-upon keyword (model, framework, etc are often used interchangeably with theory).
The modular publishing movement may ameliorate the Findability of psychological theory.
In this approach, distinct resources are individually published as citable academic output.
Such output can still be linked; for example, a FAIR theory could have an accompanying paper.
Some have sought to address limitations of theories'findability through post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and post many examples on their website.
Similarly, Borsboom and colleagues seek to establish a dictionary of psychological ``phenomena'' (which are not strictly theories, but are patterns reliably evidenced by data that theory should seek to explain).

With regard to Accessibility, publishing behind paywalls certainly has a negative impact.
Open Access publishing is a practice that increases the accessibility of all academic output, including theory.
Another factor curtailing the accessibility of theory is ambiguity;
a lack of formalization introduces a dependency on the original author for clarification.
The discourse on ``Great Man Theorizing'' touches upon the problems this introduces {[}REF Guest et al moral theory{]}.
Dependency on interpretation by the author creates a potential for gatekeeping - the author could insist that work requires their involvement, which violates checks and balances of scientific research.
Moreover, if a theory is refuted, its author could claim that the author of the refuting paper did not interpret the theory correctly.
This also relates to the problem of translation {[}REF Duhem{]}:
it is not possible to entirely formalize an idea to enable unambiguous interpretation.
Nonetheless, taking care to formalize a theory to the maximum extent possible advances Accessibility.

@@

This protocol shows how to use DAGs for causal inference, including the detection of a violation of the initial model and subsequent adaptation of the DAG. We could use that to illustrate updating FAIR theory:

\url{https://currentprotocols.onlinelibrary.wiley.com/doi/full/10.1002/cpz1.45}

We can find more examples of causal inference with DAGs in these tutorials:

\url{https://www.r-bloggers.com/2019/08/causal-inference-with-dags-in-r/}

\url{https://www.r-bloggers.com/2018/08/applications-of-dags-in-causal-inference/}

@@
